\documentclass{ximera}
\title{Circles, Transformations, Ellipses}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
Transformations
\end{abstract}
\maketitle



We are accustomed to parameterizing the motion of a beetle crawling around a circle centered at the origin at a constant speed. This means that we find functions that express the beetle's coordinates in terms of time. These functions tell us its coordinates at any time in the domain. Input a time and the functions spit out the coordinates.

There are two steps behind this type of parameterization. The first is to express the beetle's polar angle (really the polar angle of the vector from the origin to the beetle) in terms of time. The second is to use this polar angle function together with the definitions of the cosine and sine functions to express the coordinates as functions of time.

One key point, the main idea of our entire class, is that with $(x,y)$ the coordinates of the beetle and $\theta$ its polar angle, we know (by definition) that
\[
   \cos\theta = \frac{x}{r} \hskip 0.5 in \text{ and } \hskip 0.5 in \sin\theta = \frac{y}{r},
\] 
where $r = \sqrt{x^2+y^2}$ is the radius of the circular path. This tells us
\[
   x = r \cos \theta \hskip 0.5 in \text{ and } \hskip 0.5 in  y = r\sin \theta.
\]

These equations parameterize the circle of radius $r$ centered at the origin in terms of the polar angle $\theta$ of the vector $\overrightarrow{OP}$ from the origin to the beetle. Input a polar angle $\theta$ and the functions give the beetle's $x$ and $y$ coordinates.

\begin{question} \label{Q9edDEFRLERr3}
Suppose $r=10$ cm. Use the above parameterization to find the coordinates of the beetle when it has polar angle
\begin{enumerate}
\item $\theta = 0$

\item $\theta = \pi/2$

\item $\theta = \pi$.

Check that these coordinates are correct.
\end{enumerate}
\end{question}

\begin{question}  \label{Ex:sdfdsfdfL}

But what if the circular path is not centered at the origin? We address this question next.

\section{Circles Not Centered at the Origin}

We'll parameterize a circle with radius $3$ cm centered at the point $C(5,4)$ (coordinates measured in cm) in terms of the polar angle of the vector from $C$ to a point $P(x,y)$ on the circle. The steps are outlined below.

\begin{onlineOnly}
    \begin{center}
\desmos{pcndel98wg}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/pcndel98wg}{142: Translating Circles}


\begin{explanation} 

Here are the steps, with the details left for you to fill in.

We'll start by letting $P$ be a point on the circle with coordinates $(x,y)$ and $\theta$ the polar angle of the vector $\overrightarrow{CP}$ giving the position of $P$ relative to $C$.

\begin{enumerate}
\item First express the components of the vector $\overrightarrow{CP}$ in terms its polar angle $\theta$.

\item Next express the vector $\overrightarrow{OP}$ giving the position of $P$ relative to the origin in terms of $\overrightarrow{CP}$ and the vector $\overrightarrow{OC}$.

\item Use the results of parts (a) and (b) to express the components of the vector $\overrightarrow{OP}$ in terms of $\theta$.

\item Use the result of part (c) to find functions
\[
    x = f(\theta)  \hskip 0.5 in \text{and} \hskip 0.5in  y=g(\theta) ,
\]
that express the coordinates of $P$ in terms of $\theta$. Include appropriate domains for these functions. 

\item One last question. Find equations in Cartesian coordinates of 

\begin{enumerate}
\item the unit circle centered at the origin and 

\item The circle of radius $3$ centered at  $C(5,4)$.

These equations express relationships between the $x$ and $y$ coordinates of points on the circles. They do not involve the parameter $\theta$. 

\end{enumerate}
\end{enumerate}

\end{explanation}
\end{question}


\begin{question}  \label{QDF09dfdfLKDD}
\begin{enumerate}

\item Parametrize the circle in Question \ref{Ex:sdfdsfdfL} by the signed arclength from $A$ of a point $P(x,y)$ on the circle. This means to express the coordinates $(x,y)$ of $P$ (measured in cm) in terms of its signed distance from $A$. Measure the signed arclength $s$ (in cm)  from the point $A(8,0)$ along the circle and take the counterclockwise direction to be positive. 

\begin{onlineOnly}
    \begin{center}
\desmos{llrzcv5ckf}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/llrzcv5ckf}{142: Circle Arclength Parameterization}



The coordinate functions (in cm) are 
\[
  x = f_1(s) = \answer{5 + 3 \cos(s/3)} \, , \, s\in \mathbb{R}
\]
and
\[
   y = g_1(s) = \answer{4 + 3 \sin(s/3)} \, , \, s\in \mathbb{R} .
\]

\item Use part (a) to find the exact coordinates of the point on the circle  $20$ cm from $A$ as measured counterclockwise along the circle. 

\item Use the worksheet above to approximate these coordinates and check if your coordinates look reasonable.

\item Between noon and 12:02pm a beetle crawls counterclockwise around the circle at a constant speed of $6$cm/sec, passing the point $A(8,0)$ at $23$ seconds past noon. Find coordinate functions
\[
   x = f_2(t)  \hskip 0.5 in \text{and} \hskip 0.5in  y=g_2(t) ,
\] 
that express the beetle's coordinates (in cm) in terms of the number of seconds past noon.

\end{enumerate}
\end{question}

\section{Ellipses}

Look at a soccer ball on a sunny day and you'll see an ellipse. It's the boundary of the ball's shadow. Or cut a circlular cylinder with a plane. A horizontal plane gives a circle, but tilting the plane gives an ellipse.

This section explores two ways to parameterize an ellipse. We first need to know what an ellipse is. One way to get an ellipse is by dilating (ie. stretching or compressing) a circle about one of its diameters. To make things simple, we'll start with the unit circle (radius one, centered at the origin). To see an example of this stretching action, first drag the slider $k_1$ below from $k_1=1$ to $k_1=4$. Then drag the slider $k_2$ from $k_2=1$ to $k_2=2$.

\begin{onlineOnly}
    \begin{center}
\desmos{l2wiiamr0a}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/l2wiiamr0a}{142: Ellipse 1B}


\begin{question} \label{QLDkferedsf}
\begin{enumerate}
\item Describe the effects of these stretching actions geometrically.

\item Describe how the two stretches change the coordinates of a point on the circle.
\end{enumerate}
\end{question}



\begin{example} \label{Ex:DFDrefdfobbzx}
Our goal now is to parameterize the above ellipse. We'll start with the standard parameterization 
\[
    (x,y) = (f(\theta) , g(\theta)) =   ( \cos \theta , \sin \theta ) \, , \, 0\leq\theta < 2\pi ,
\]
of the unit circle centered at the origin. Here $\theta$ is the polar angle of the vector $\overrightarrow{OP}$ from the origin to a point $P(x,y)$ on the circle.

Now replace $x$ with $x/4$ and $y$ with $y/2$ to get a parameterization
\[
  \frac{x}{4} = f(\theta) = \cos\theta \, , \, 0\leq\theta < 2\pi ,
\]
\[
  \frac{y}{2} = g(\theta) = \sin \theta \, , \, 0\leq\theta < 2\pi 
\]
of a new curve. How is this new curve related to the unit circle?

We might start by rewriting the new parameterization as
\[
      x = 4 f(\theta) = 4\cos\theta \, , \, 0\leq\theta < 2\pi ,
\]
\[ 
   y = 2 g(\theta) = 2\sin \theta \, , \, 0\leq\theta < 2\pi .
\]

What this means is that to get the coordinates of points on the new curve we quadruple the $x$-coordinates and double the $y$-coordinates of the points on the unit circle. Geometrically, this has the effect of dilating (stretching) the unit circle horizontally by a factor of four about the $y$-axis and vertically by a factor of two about the $x$-axis. 


\begin{onlineOnly}
    \begin{center}
\desmos{3ibj708prx}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/3ibj708prx}{142: Ellipse 1}

\begin{enumerate}
\item Watch these stretching actions by dragging the dilation factors $k_1$ and $k_2$ in the worksheet above from $k_1=1$ to $k_1=4$ and $k_2=1$ to $k_2 = 2$ in the worksheet above.

\item If the point $P$ on the unit circle has coordinates $(a,b)$, what are the coordinates of its image $P^\prime$ after these dilations?

After the two dilations $P^\prime$ has coordinates
\[
    (a^\prime, b^\prime) = (  \answer{4a} , \answer{2b} ) .
\]

\item Find an equation in Cartesian coordinates of the new curve (an ellipse). This equation gives a relationship between the $x$ and $y$ coordinates of a point $P(x,y)$ on the ellipse. It does not involve the parameter $\theta$.

\item Find the exact coordinates of the point $P^\prime$ on the ellipse with parameter $\theta=\pi/4$. Then use a calculator to approximate these coordinates and the worksheet above as a check (drag the slider $u$ on Line 6 to $u=\pi/4 \sim 0.785$).

\item Find the exact coordinates of the point $Q$ on the ellipse with polar angle $\pi/4$. Then find the exact polar radius (ie. the distance between this point and the origin). Use the worksheet below to check your coordinates.

\begin{onlineOnly}
    \begin{center}
\desmos{4ayj4l4hca}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/4ayj4l4hca}{142: Ellipse 1C}





\item Working from the parameterization 
\[ 
     x = f_2 (\theta) = 4\cos\theta \, , \, 0\leq \theta <2\pi ,
\]
\[
   y = g_2 (\theta) = 2\sin\theta \, , \, 0\leq \theta <2\pi ,
\]
of the ellipse, let's replace $x$ with $x-5$ and $y$ with $y-4$ to get a parameterization of a new curve.

\begin{enumerate}
\item Sketch by hand the new curve.

\item Find an equation of the new curve in Cartesian coordinates.

\item Remember our point $P$ with coordinates $(a,b)$ on the unit circle. Replacing $x$ with $x/4$ and $y$ with $y/2$ in the parameterization of the unit circle sent the point $P$ to the point $P^\prime$ with coordinates $(4a, 2b)$ on the ellipse
\[
  \left( \frac{x}{4} \right)^2 + \left( \frac{y}{2} \right)^2 = 1.
\]
Now replacing $x$ wiht $x-5$ and $y$ with $y-4$ gives us a new ellipse with equation
\[
  \left( \frac{x-5}{4} \right)^2 + \left( \frac{y-4}{2} \right)^2 = 1
\]
and sends $P^\prime(4a,2b)$ to a new point $P^{\prime\prime}$ on the new ellipse. What are the coordinates of $P^{\prime\prime}$?

The coordinates are 
\[
   (a^{\prime\prime} , b^{\prime\prime}) = (\answer{4a+5} , \answer{2b+4} ) .
\]


\end{enumerate}

\end{enumerate}
\end{example}

\begin{example} \label{EX:9dfrebbvcxbfg}
Return now to the ellipse with paramterization
\[
      (x,y) = (4\cos \theta , 2\sin\theta) \, , \, 0\leq \theta <2\pi ,
\]
and Cartesian equation
\[
     \left( \frac{x}{4} \right)^2 + \left( \frac{y}{2} \right)^2 = 1.
\]

While $\theta$ is the polar angle of the point $(\cos\theta, \sin\theta)$ on the unit circle, it is \emph{not} the polar angle of the point $(4\cos \theta , 2\sin\theta)$ on the ellipse. We saw this in the previous question where the parameter $\theta=\pi/4$ was not the polar angle of the point $P^\prime(4\cos (\pi/4) , 2\sin (\pi/4))$ on the ellipse.

%But just to make sure you understand this, drag the slider $u$ on Line 2 of the worksheet below to convince yourself that the parameter $\theta$ in the parameterization 
%\[
%    (x,y) = (4\cos \theta , 2\sin\theta ) \, , \, 0 \leq \theta <2\pi 
%\]
%of the ellipse is \emph{not} the polar angle of the vector $\overrightarrow{OP^\prime}$. Explain how you know.

%\begin{freeResponse}
%\end{freeResponse}


\begin{onlineOnly}
    \begin{center}
\desmos{3skcbdibwx}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/3skcbdibwx}{142: Ellipse 2B}

Our aim in this problem is to parameterize this ellipse by its polar angle $\theta$.  It is the polar angle of the vector $\overrightarrow{OQ}$ from the origin to the point $Q$ on the ellipse shown above. We wish to express the coordinates $(x,y)$ of $Q$ in terms of $\theta$.


\begin{enumerate}

\item To parameterize the ellipse in terms of the polar angle $\theta$ means to express the $x$ and $y$ coordinates of $Q$ in terms of $\theta$. To do this, we need to first express the polar radius  
\[
   r = \Big| \overrightarrow{OP^\prime}  \Big|
\]
in terms of $\theta$. Here are the steps:

%\begin{onlineOnly}
%    \begin{center}
%\desmos{le8nhq7iwq}{900}{600}
%\end{center}
%\end{onlineOnly}

%\href{https://www.desmos.com/calculator/le8nhq7iwq}{142: Ellipse 3}


\begin{enumerate}
\item First express the coordinates of $P^\prime$ in terms of $r$ and $\theta$.

\item Then use algebra and the Cartesian equation of the ellipse to find an equation relating $r$ and $\theta$.

\item Next solve this equation for $r$ in terms of $\theta$ keeping in mind that $r>0$.

The result is that
\[
  r = \answer{\frac{8}{\sqrt{4\cos^2\theta + 16\sin^2\theta }}} .
\]

\item Finally, use the results of parts (i) and (iii) to parameterize the ellipse in terms of the polar angle $\theta$.
\end{enumerate}

\end{enumerate}
\end{example}

\begin{example} \label{Ex:8UDUREGDFGD}
Between noon and 12:30pm a beetle travels counterclockwise around the ellipse
\[
         \left(  \frac{x}{4} \right)^2 + \left(  \frac{y}{2} \right)^2  = 1 ,
\]
passing the point $(4,0)$ at 12:23pm (coordinates measured in meters). The beetle rotates about the origin at a constant rate, making one revolution every 5 minutes. 

\begin{enumerate}

\item Find functions
\[
    x = f(t)  \hskip 0.5 in \text{and} \hskip 0.5 in y=g(t) \, , \, 0\leq t \leq 30 ,
\]
that express the coordinates of the beetle (in meters) in terms of the number of minutes past noon.

\item Enter your coordinate functions in Lines 2 and 3 of the worksheet below. Then play the slider $u$ (another name for $t$) to see if you are correct.

\begin{onlineOnly}
    \begin{center}
\desmos{goeo2vvdvi}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/goeo2vvdvi}{142: Motion around an ellipse}




\end{enumerate}
\end{example}

\end{document}