\documentclass{ximera}
\title{Mars and the Outer Planets}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
The retrograde motion of Mars and the outer planets.
\end{abstract}
\maketitle

\begin{exploration} \label{Exphjhhkjglvg}

\begin{center}
\youtube{https://www.youtube.com/watch?v=1nVSzzYCAYk}
\end{center}

\href{https://www.youtube.com/watch?v=1nVSzzYCAYk}{Retrograde Motion of Mars}

\end{exploration}

\begin{exploration}  \label{Eddfsdfd6yhh}
Assume for this problem that Earth and Mars rotate about the sun at constant rates in circular orbits that lie in the same plane. Mars has an orbital period of Mars to be 687 earth days and an orbital radius of $1.524$ astronomical units. Earth has an orbital period of $365$ days and an orbital radius of $1$ astronomical unit.

\begin{enumerate}

\item Parameterize the motions of earth and mars about the sun in terms of the number of earth days. Assume

\begin{itemize}

\item{The sun is at the origin.}

\item{The planets rotate counterclockwise at constant rates about the sun.} 

\item{The earth has coordinates $(1,0)$ a time $t=0$ days.}

\item{The vector from the sun to Mars has polar angle $\theta_0$ at time $t=0$.}

\end{itemize}

\item Express the position of Mars relative to Earth in terms of the vectors $\overrightarrow{SE}$ and $\overrightarrow{SM}$
that respectively give the positions of Earth and Mars relative to the sun.


\item Use part (b)  to parameterize the motion of mars relative to the earth (with the earth  fixed at the point $(0,-4)$.


\item Check your work using the demonstration below by inputting the correct functions in Lines 3, 4, 6, 7, 9, and 10.

Access Desmos interactive at
 
\href{https://www.desmos.com/calculator/htc4xgrjxs}{Mars Retrograde Motion}

 
\begin{onlineOnly}
    \begin{center}
\desmos{htc4xgrjxs}{900}{600}
\end{center}
\end{onlineOnly}

\item Compute the synodic period of Mars as seen from Earth. This is the time, measured in earth days, it takes Earth to ``lap'' Mars. It is the time it takes Mars to return to the same position relative to the sun as seen from Earth. 



\end{enumerate}

Access Geogebra interactive at
 
\href{https://www.geogebra.org/classic/addm38j6}{Mars Retrograde Motion}

 
\begin{onlineOnly}
    \begin{center}
\geogebra{addm38j6}{900}{600}
\end{center}
\end{onlineOnly}



\end{exploration}




\end{document}