\documentclass{ximera}
\title{Vectors}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
An introduction to vectors in the plane.
\end{abstract}
\maketitle

\section{Position Vectors}

Position is relative. You describe where something is not in absolute terms but only in relation to something else.

We might say, for example, that Bellevue is 4 miles east and 3 miles north of Akron. Then in a rectangular coordinate system with the $x$-axis pointing due east, the $y$-axis pointing due north, and the coordinates measured in miles, we can represent the position of Bellevue (at point $B$) relative to Akron (at point $A$) by the vector
\[
   \overrightarrow{AB} = \langle 4, 3 \rangle .
\]
This vector tells how to get from Akron to Bellevue. 


Suppose that in our rectangular coordinate sytstem Charleston (point $C$) has coordinates $(-5,-2)$ and Dayton (point $D$) has coordinates $(-1,1)$. Then to get from Charleston to Dayton we would travel east
\[
     ( -1 - (-5) ) \text{ miles} = 4 \text{ miles}
\]
and north
\[
     ( 1 - (-2)) \text{ miles} = 3\text{ miles} .
\]
So we can represent the position of Dayton relative to Charleston by the vector
\[
  \overrightarrow{CD} =       \langle -1 - (-5) , 1- (-2)  \rangle   =  \langle 4, 3 \rangle .
\] 

So the directions that tell us how to get from Akron to Bellevue also tell us how to get from Charleston to Dayton. The
vectors $\overrightarrow{AB}$ and $\overrightarrow{CD}$ are equal and we write 
\[
      \overrightarrow{AB} = \overrightarrow{CD} .
\]

To go the other way and get from Bellevue to Akron or from Dayton to Charleston, we would go 4 miles west and 3 miles south. So the  position of Akron relative to Bellevue, or of Charleston relative to Dayton, is given by the vector
\[ 
   \overrightarrow{BA} = - \overrightarrow{AB} = - \langle 4,3 \rangle =  \langle -4,-3 \rangle.
\]

The distance between Akron and Bellvue is just the length of the vector $\overrightarrow{AB}$, written as
\[
    | \overrightarrow{AB}  | =  | \langle 3, 4 \rangle  | \text{ miles} = \sqrt{3^2 + 4^2} \text{miles} = 5 \text{ miles} .
\] 


\section{Adding and Subtracting Vectors}

\begin{question}    \label{Q234234:Vectors}
Continuing with our example above, suppose that Frankfort (point $F$) is $2$ miles east and $5$ miles south of Bellevue. Then the vector
\[
    \overrightarrow{BF}  = \langle \answer{2}, \answer{-5}\rangle
\]
gives the position of Frankfort relative to Bellvue. 
\end{question}

\begin{question} \label{Q233:Vectors}
(a) Give directions to get from Akron to Frankfort.

(b) Fill in the vector and its components that gives the position of Franfort relative to Akron.
\[
       \overrightarrow{\answer{A} \answer{F}}    =  \langle  \answer{6} , \answer{-2}  \rangle 
\]

(c) Which of the following are correct? Choose all that apply.
\begin{selectAll}  
\choice[correct]{$\overrightarrow{AF}   = \overrightarrow{AB} + \overrightarrow{BF}$}  
\choice{$\overrightarrow{AF}   = \overrightarrow{AB} - \overrightarrow{BF} $}  
\choice[correct]{$\overrightarrow{AF}   = \overrightarrow{BF} + \overrightarrow{AB} $}  
\end{selectAll} 


\end{question}


\begin{question}  \label{Q543:Vectors}
Let ${\bf v}$ be the vector that gives the position of point $K$ relative to point $G$ and ${\bf w}$ the vector that gives the position of $M$ relative to $K$. 

(a) Express in terms of ${\bf v}$ and ${\bf w}$ each of the following vectors: 

(i) The position of $M$ relative to $G$:  $\overrightarrow{\bf \answer{v}} +  \overrightarrow{\bf \answer{w}} $

(ii) The position of $G$ relative to $M$:  $\overrightarrow{\bf \answer{-v}} +  \overrightarrow{\bf \answer{-w}} $

\pskip

(b) Draw vectors ${\bf v}$ and ${\bf w}$ tail-to-tail. Any choices will do, just make sure the vectors are not parallel. Then draw the vector ${\bf v} + {\bf w}$

(c) Explain the meaning of each of the following in the context of this scenario. Include units. 

(i) $|  {\bf v} |$

(ii) $|  {\bf w} |$

(iii) $|  {\bf v+w} |$

(iv) $|  {\bf v} | + |  {\bf w} | - |  {\bf v+w} |$

\pskip

(d) When would the expression in part (b)(iv) above be positive? Negative? Zero? Explain your reasoning. 

\end{question}


\begin{question}  \label{Qsdfsdt4r3:Vectors}
Suppose
\[
 \overrightarrow{AB} = \langle  -3,5 \rangle
\]
and
\[
   \overrightarrow{AC} = \langle  4,2 \rangle ,
\]
where the components are measured in miles.

(a) Give directions to get from $B$ to $C$.

(b) What vector gives the position of $C$ relative to $B$? Choose all that apply.
\begin{multipleChoice} 
 \choice[correct]{$\langle  7, -3  \rangle$}  
 \choice{$\langle  -7, 3  \rangle$}
\choice{$\overrightarrow{CB}$}  
\choice[correct]{$\overrightarrow{BC}$} 
\choice{$\overrightarrow{AB} -\overrightarrow{AC} $} 
\choice[correct]{$\overrightarrow{AC} -\overrightarrow{AB} $}
\end{multipleChoice} 

\end{question}





\begin{question}  \label{Q234r3:Vectors}
Let ${\bf v}$ be the vector that gives the position of point $Q$ relative to point $R$ and ${\bf w}$ the vector that gives the position of $S$ relative to $R$. 

(a) Which of the following express the position of $S$ relative to $Q$? Choose all that apply.

\begin{selectAll}  
\choice{${\bf v} + {\bf w}$}  
\choice{${\bf v} - {\bf w}$}  
\choice[correct]{${\bf w} - {\bf v}$}  
\end{selectAll}

(b) Draw a picture illustrating part (a). Include the vectors ${\bf v}$ and ${\bf w}$ tail-to-tail, and the vector that gives the position of $S$ relative to $R$. That's it. Do not draw a fourth vector like $-{\bf v}$ or $-{\bf w}$.


\end{question}



IGNORE THIS

\pdfOnly{
Access Desmos interactives through the online version of this text at
 
\href{https://www.www.geogebra.org/classic/e6rsvnsz}.
}
 
\begin{onlineOnly}
    \begin{center}
\geogebra{e6rsvnsz}{900}{600}
\end{center}
\end{onlineOnly}



\section{Computing Coordinates With Vectors}

\begin{question} \label{Q45340:Vectors}
(a) Explain why the circles with equations
\[
   (x-2)^2 + (y+1)^2  = 49
\]
and
\[
    (x-5)^2 + (y-3)^2 = 4
\]
are tangent to each other.

(b) Use vector arithmetic, not algebra, to determine the coordinates of the point of tangency.


\end{question}


\begin{question} \label{Q45323240:Vectors}
Points $A$ and $B$ have respective coordinates $(5,1)$ and $(17,-15)$. 

(a) Use vectors, not algebra, to determine the coordinates of the point $P$ on segment $\overline{AB}$ that is exactly 17 units from $A$.

(b) Use vectors, not algebra, to determine the coordinates of another point $P$ on the line through $A$ and $B$ that is exactly 17 units from $A$.

(c) Solve questions (a) and (b) simultaneously using algebra, not vectors.

\end{question}


\begin{question} \label{Q4df8240:Vectors}
(a) Use vector arithmetic, not algebra, to determine the coordinates of the centers of both circles with radius 6 that are tangent to the line
\[
    2x + 5y = 13
\]
at the point $P(-1,3)$.

(b) Use the method of part (a) to determine the coordinates of the centers of all circles with radius $r$ that are tangent to the same line at the same point.

(c) Follow the directions in the demonstration below to check your work.

\pdfOnly{
Access Desmos interactives through the online version of this text at
 
\href{https://www.desmos.com/calculator/aifh0gprxt}.
}
 
\begin{onlineOnly}
    \begin{center}
\desmos{aifh0gprxt}{900}{600}
\end{center}
\end{onlineOnly}

\end{question}


\begin{question}   \label{Q6547:Vectors}
(a) Use vectors, not trigonometry, to help find equations of the two lines that bisect the angles between the lines $y=x$ and $y=3x$.

(b) Use part (a) to find equations of the two lines that bisect the angles between the lines $y=x+3$ and $y=3x-5$.


\end{question}


\section{Position Vectors and Bearing}

Bearings are typically measured clockwise from due north. So you might report a fire to be $10$ miles from Miner's Ridge Lookout at a bearing of $100^\circ$. But we'll continue to meaure our angles counterclockwise (in radians) from the direction of the positive $x$-axis. We'll use polar angle and bearing interchangeably to mean the same thing.

\begin{example}  \label{Edsa5454v:Vectors}
Point $B$ is 4 miles from $A$ at a bearing of $3$ radians. Point $C$ is $7$ miles from $B$ at a bearing of $4.1$ radians. We wish to describe the position of $C$ relative to $B$ in two ways:

i) By giving north-south-east direction. For example, by saying that $C$ is $2$ miles east and $6$ miles south of $A$. We'll call this the NESW description.

ii) By giving a distance and a bearing. For example by saying that $C$ is $5$ miles from $A$ at a bearing (polar angle) of $5.3$ radians. We'll call this the {\bf distance-bearing} description.

Neither of these descriptions is correct for our problem. But we'll now determine the correct directions.

\begin{explanation}

(a) We'll start by using the demonstration below to draw a reasonably accurate picture. %Use the radian protractor and varying its radius $r$ with the slider and checking the correct boxes for vectors, we get the following picture

\pdfOnly{
Access Desmos interactives through the online version of this text at
 
\href{https://www.geogebra.org/classic/bhdsgxtx}.
}
 
\begin{onlineOnly}
    \begin{center}
\geogebra{bhdsgxtx}{900}{600}
\end{center}
\end{onlineOnly}


(b) We are given enough information about the vector $\overrightarrow{AB}$ to find its components (measured in miles) in the rectangular coordinate system above, where the unlabeled, positive $x$-axis points due east and the positive $y$-axis due north.

Because the distance from $A$ to $B$ is $r=  |\overrightarrow{AB}| = 4$ miles and $\overrightarrow{AB}$ is inclined at an angle of $3$ radians relative to the positive $x$-axis, the vector
\[
       \overrightarrow{AB} = \langle 4 \cos 3, 4 \sin 3\rangle .
\]
gives the position of $B$ relative to $A$. Its components are measured in miles.

Similarly, the vector
\[
   \overrightarrow{BC} = \langle 7 \cos 4.2, 7 \sin 4.1\rangle
\]
gives the position of $C$ relative to $B$ (in miles).

To find the position of $C$ relative to $A$, we add these vectors. So the position of $C$ relative to $A$ (in miles) is 
\begin{align*}
\overrightarrow{AC} &= \overrightarrow{AB} + \overrightarrow{BC} \\
                              &= \langle 4 \cos 3, 4 \sin 3\rangle + \langle 7 \cos 4.1, 7 \sin 4.1\rangle \\
                              & =    \langle 4 \cos 3 + 7 \cos 4.1 , 4 \sin 3 + 7 \sin  4.1\rangle   \\
                              & \sim \langle  -7.98 ,  -5.16  \rangle 
\end{align*}

So to get from point $A$ to point $C$ you would walk
\[
     (4 \cos 3 + 7 \cos 4.1) \text{  miles} \sim 7.98 \text{ miles} 
\]
due west and 
\[
   (4 \sin 3 + 7 \sin 4.1) \text{  miles} \sim 5.16 \text{ miles} 
\]
due south.

\pskip

(c) For the distance-bearing description of how to get from $A$ to $C$, we'll fiirst compute the exact distance between $A$ and $C$. This distance in miles is
\begin{align*}
   | \overrightarrow{AB}| &= |   \langle 4 \cos 3 + 7 \cos 4.1 , 4 \sin 3 + 7 \sin  4.1\rangle   |   \\ \\
                                   & = \sqrt{( 4 \cos 3 + 7 \cos 4.1)^2 + (4 \sin 3 + 7 \sin  4.1)^2}  \\ \\
                                  & =  \sqrt{16 + 49 + 56 (\cos(3) \cos(4.1) + \sin(3)\sin(4.1))}  \\ \\
                                  & \sim 9.46
\end{align*}

We do not know enough yet to get the exact bearing from $A$ to $C$ (we'll need to learn about inverse trig first), but for now we'll just use the radian protractor to estimate the bearing. 

Let's drag the protractor's center to $A$ and first check the distance between $A$ and $C$ by dragging the slider $r$ that controls the protractor's radius. Check that $r\sim 9.4$. Then read off the polar angle of the vector $\overrightarrow{AC}$ from the protractor. 

\begin{question}  \label{Q43656ff:Motion}
The bearing from $A$ to $C$ is about 
\begin{multipleChoice}  
\choice{$3$ radians}  
\choice[correct]{$3.7$ radians}  
\choice{$0.7$ radians}  
\end{multipleChoice} 

\end{question} 


So to get directly from $A$ to $C$ we should walk about $9.5$ miles at a bearing of approximately $3.7$ radians (actually closer to $3.6$ radians).

\end{explanation}


\end{example}


\begin{question} \label{Qdgg4tnh:Vectors}
Point $B$ is 5 miles from $A$ at a bearing of $2$ radians. Point $C$ is $8$ miles from $A$ at a bearing of $5$ radians. Describe  the position of $C$ relative to $B$ in the two ways of the previous example - the NESW description (ie. the component form of a vector) and the distance-bearing description (the polar form of a vector). Follow all the steps in the previous example, making modifications as necessary.

\end{question}


\begin{question}  \label{Qgtsygh:Vectors}
Starting from the point $A$ with coordinates $(-4 \text{ miles},-2 \text{ miles})$, you walk $8$ miles directly toward the point $Q(6,8)$ and stop at point $B$. Still facing point $Q$, you then turn clockwise through an angle of $7\pi/12$ radians, walk $10$ miles, and stop at point $C$. Describe the position of the origin relative to point $C$ in the same two ways as before. Follow all the steps in the previous example, making modifications as necessary. 

\end{question}



\end{document}