\documentclass{ximera}
\title{Right Triangle Trigonometry, Part 2}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
More right triangle tirg.
\end{abstract}
\maketitle


\section{Describing Position on the Earth}

The \emph{latitude} of a point $P$ on the surface of the earth is the angle the radius $OP$ from the earth's center to $P$ makes with the plane of the equator. Latitude, often represented by the Greek letter $\phi$ (phi), lies between $-\pi/2$ (at the south pole) and $\pi/2$ at the north pole. It is positive in the northern hemisphere, negative in the southern hemisphere.

Click on the circular icon to the left of Line 2 in the worksheet below to hide the earth. The two marked angles measure the latitude of the blue point (unlabeled) on the sphere.

\begin{onlineOnly}
    \begin{center}
\desmosThreeD{pio4iubhvt}{800}{600}         %zk06s3k6q4
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/3d/pio4iubhvt}{142: Circle of Latitude}


\begin{question} \label{Qfksadfsdt4e4}
\begin{enumerate}
\item How fast are we moving due to the rotation of the earth about its axis? Assume Shoreline is at a latitude of $47.75^\circ$N and take the radius of the earth to be $4000$ miles. Find the exact speed \emph{without} a calculator. Then use a calculator to approximate the speed to the nearest mile/hour.

{\bf Do not rely on a formula.} Use logic and common sense instead. You will need to know the definition of $\pi$. Can you remember this from the first day of class?

\item At what latitude(s) are people moving at a speed of $200$ miles/hour due to the rotation of the earth about its axis? Find the exact latitude(s). Then use a calculator to approximate their radian and degree measures.

Start by defining your unknown. Then write an equation and solve for the unknown.
\end{enumerate}
\end{question}

\begin{question}  \label{QODfefEREr}
Assume for this problem that Shoreline and Spokane are both at a latitude of $47.75^\circ$N. Suppose also that the cities have respective longitudes $122.3^\circ$W and $117.4^\circ$W. 

Click the camera icon at the left of Line 1 to hide the sphere.

\begin{onlineOnly}
    \begin{center}
\desmosThreeD{rjb4gcf8lo}{800}{600}         %zk06s3k6q4
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/3d/rjb4gcf8lo}{142: Seattle to Spokane}


\begin{enumerate}
\item Find the distance between the two cities measured along their common circle of latitude.

\item Compare this distance with the driving distance along I-90, about $270$ miles.

\end{enumerate}

\end{question}


\section{Other Questions}

\begin{question} \label{QdfRERER}
You wrap a rubber band around a jar lid of radius $R$ cm and stretch the band as shown below.

Express the length (measured in cm) of the band in terms of $R$ and the distance $h$ between the point $P$ and the center of the lid, also measured in cm.


\begin{onlineOnly}
    \begin{center}
\desmos{wz5qiyj4od}{800}{600}         %zk06s3k6q4
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/wz5qiyj4od}{142: Jar Lid}

\end{question}


\begin{question} \label{QddtgHHERER}
Sensors on the ground at points $A$ and $B$ below measure the angles of elevation to a plane to be $\theta_1$ and $\theta_2$, respectively, as shown below.

Express the altitdue of the plane (in km) in terms of the distance $s$ between the sensors (also measured in km).

\begin{onlineOnly}
    \begin{center}
\desmos{zbktegkke4}{800}{600}         %zk06s3k6q4
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/zbktegkke4}{142: Helicopter}

\end{question}


\begin{question}  \label{QQQ343f44fd}
At noon a rowboat is $10$ km due east of a sailboat. The sailboat travels at a consant speed of $v$ km/hour at a fixed bearing of $0.4$ radians (measured counterclockwise from the east). The rowboat travels half as fast as the sailboat at some fixed bearing. Sometime later the boats collide.

(a) Find the possible bearing(s) of the rowboat.

(b) When do the boats collide?

(c) Establish a rectangular coordinate system with the positive $x$-axis pointing due east, and the origin at the sailboat's position at noon. Then parameterize the motions of the rowboat and sailboat. Consider all possibilities. Include domains and do \emph{not} use inverse trigonometric functions in the parameterizations.

(d) Follow the directions in the Desmos acitivity below to check your work.

\pdfOnly{
Access Desmos interactives through the online version of this text at
 
\href{https://www.desmos.com/calculator/xjiwflhe2o}.
}
 
\begin{onlineOnly}
    \begin{center}
\desmos{xjiwflhe2o}{900}{600}
\end{center}
\end{onlineOnly}

Access Desmos interactive at

\href{https://www.desmos.com/calculator/xjiwflhe2o}{142: Sailboat and Rowboat}

\end{question}


\end{document}