\documentclass{ximera}
\title{Inclined Cycloids}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
Inclinded cycloids.
\end{abstract}
\maketitle

\begin{exploration}  \label{Exhjjjjnbbbbbr}

A wheel of radius $R$ meters rolls without slipping on the $x$-axis as a pebble $P$ attached to the wheel traces out a curve (drag the slider $u$ in the demonstration below). The wheel's axle $C$ moves  at a constant speed of $v$ meters/sec. Our aim is to parameterize the motion of the pebble.


\href{https://www.geogebra.org/classic/vxbjccpa}{Cycloid}

 
\begin{onlineOnly}
    \begin{center}
\geogebra{vxbjccpa}{900}{600}
\end{center}
\end{onlineOnly}


\begin{enumerate}
\item Express the components of the vector $\overrightarrow{OC}$ in terms of the rotation angle $\theta$ of the wheel, taking $\theta=0$ when the pebble is at the origin.

\item Turn on the \emph{axes} box in the upper left corner of the workseet. Then express the components of the vector $\overrightarrow{CP}$ in terms of $\theta$.

\item Express the position of $P$ relative to the origin in terms of the vectors $\overrightarrow{OC}$ and $\overrightarrow{CP}$.

\item Use part (c) to express the components of the vector $\overrightarrow{OP}$ in terms of $\theta$.

\item Find a function 
\[
   \theta = h(t)  \, , \, t\geq 0, 
\]
that expresses the rotation angle in terms of $t$ (the number of seconds since the pebble left the origin).

\item Express the components of $\overrightarrow{OP}$ in terms of $t$.

\item Find coordinate functions
\[
    x=f(t)   \hskip 0.5in \text{and} \hskip 0.5in y=g(t) \, , \, t\geq 0 ,
\]
that express the coordinates of $P$ in terms of $t$.
\end{enumerate}



the number of seconds $t$ since the the pebble left the origin.

\end{exploration}

\end{document}